``Interestingness'' in Mathematical Theory Exploration

### 2018-07-05 PhDThesis/benchmark.tex
Thesis: Bring in some changes from TEB

Our main contribution is a benchmarking methodology, shown in
Figure~\ref{figure:flow_chart} on page~\pageref{figure:flow_chart}, which both
generates a large definition/ground-truth corpus, and provides a scalable,
statistical approach to evaluating MTE tools using this corpus. We follow a
precision/recall approach similar to prior work, with the main difference being
the source of definitions and ground truths: we take existing problem sets
designed for automated theorem proving, and adapt their content for use in the
conjecture generation setting.

@@

\begin{itemize}
\item For each problem, there should be a clear distinction between the
  statement to be proved and the definitions involved, such that the two can be
  easily and meaningfully separated. This rules out problem sets like those of
  SMT-COMP~\cite{barrett2005smt}, where many problems involve uninterpreted
  functions, whose behaviour is \emph{implicit} in the logical structure of the
  theorem statement but not separately \emph{defined}.
\item Definitions should be translatable into a form suitable for the MTE tools
  under study. Our application in $\S$\ref{sec:application} requires Haskell and
  Isabelle translations, and also benefits from having definitions be strongly
  typed.
\item The problem set should be relevant to the desired domain. Our focus on
  functional programming requires higher-order functions and inductively defined
  datatypes, which rules out first-order languages/logics (such as
  TPTP~\cite{sutcliffe2009tptp}).
\item The problem set should ideally contain \emph{every} ``interesting''
  property involving its included definitions, since non-membership in the
  ground truth will be treated as being ``uninteresting''. More realistically,
  we should aim for each definition to have \emph{multiple} properties; rather
  than a mixture of unrelated problems, cherry-picked from different domains.
\item Larger problem set are preferable as they give more robust statistics, all
  else being equal (i.e. when this does not sacrifice quality).
\end{itemize}

@@

We can summarise runtimes by choosing their median, as this is more robust than
the mean against long-running outliers, and hence represents performance for a
``typical'' input of that size. Unlike previous one-off evaluations, we can also
compute the \emph{spread} of our data, for example with the inter-quartile
range.

The precision and recall for multiple samples can be combined in two ways. We
can calculate their mean value, known as the \emph{average of the ratios}, to
find the expected precision and recall for runs of this size. Alternatively, we
can sum their numerators and denominators to get the \emph{ratio of the
  averages}: this normalises the output and ground truth cardinalities, allowing
precision and recall to be modelled as simple Bernoulli processes.

We prefer the latter, where each generated conjecture (wanted or not) is like a
coin toss whose bias is the precision; likewise, the ground truth properties
(found or not) are like coin tosses biased by the recall. Note that this simple
model does not explain how many conjectures the tool decides to output for each
run.

@@

To measure progress and encourage competition in the field, it is important to
compare the relative performance of different tools on the same task. Since the
aggregate statistics in the above summaries have obscured the details of
specific runs, any comparison based on them (such as comparing average recall)
would have very low statistical power. More direct comparisons can be made using
\emph{paired} tests, since for each individual sample we have measurements for
\emph{both} tools.

We propose that running times are compared using a Wilcoxon signed-rank
test~\cite{wilcoxon1945individual}, since it does not require that the data be
normally distributed. In $\S$\ref{sec:application} we follow a variant of the
Speedup-Test protocol of Touati, Worms and Briais~\cite{touati2013speedup}; the
differences are our use of the paired form of the Wilcoxon test, and the fact
that each run was performed on a different sample (to reduce sampling bias, at
the expense of accuracy). This paired difference test is also robust to the
censoring caused by timeouts: upper-bounding the run time gives an upper-bound
to the observable difference, which biases our results in a conservative way
(\emph{towards} the null hypothesis of indistinguishable performance).

For simplicity, we suggest comparing quality by pooling together results from
all samples where both tools finished successfully. Recall can be compared using
the standard McNemar test for paired samples~\cite{mcnemar1947note}: each
ground-truth property is categorised as either being found only by the first
tool, only by the second, by both or by neither. Those found by both or neither
are ignored, and the tool-specific numbers are tested for marginal homogeneity
(i.e. whether their differences give rise to similar totals).

We cannot use McNemar's test to compare precision, since each tool may generate
different conjectures. Instead, we can use Boschloo's form of Barnard's
test~\cite{lydersen2009recommended} to compare the proportions of wanted and
unwanted conjectures generated by each tool, assuming a binomial distribution
and conditioning on the number of generated conjectures (as if this were fixed
by design).

@@

TIP comes with software to translate its definitions into Haskell and Isabelle
code, including comparison functions and random data generators suitable for
QuickCheck. We translate all 269 unique definitions into a single module/theory
which is imported on each run of the tools, although only those functions which
appear in the current sample are included in the signature and explored.

@@

\begin{figure}
  \centering
  \input{quickspectime.pgf}
  \caption{Running times of QuickSpec on theories sampled from TIP. Each point
    is an individual run, with 31 per size except for 1 (14) and 2 (30). Failing
    processes (timeouts or out-of-memory) are shown in red. Points are spread
    out horizontally to reduce overlapping. Box plots show median and quartiles,
    with whiskers showing points at 1.5$\times$~inter-quartile range.}
  \label{figure:quickspec_runtimes}
\end{figure}

\begin{figure}
  \centering
  \input{quickspecprec.pgf}
  \caption{Precision and recall of successful QuickSpec runs (spread
    horizontally to reduce overlapping). Lines show the combined proportion for
    each sample size (ratio of averages) and errorbars show the sample standard
    deviation.}
  \label{figure:quickspec_precRec}
\end{figure}

@@

\begin{figure}
  \centering
  \input{isacosytime.pgf}
  \caption{Running times of IsaCoSy on theories sampled from TIP. Each point
    is an individual run, with 31 per size except for 1 (14) and 2 (30). Failing
    processes (timeouts and out of memory) are shown in red. Points are spread
    out horizontally to reduce overlapping. Box plots show median and quartiles,
    with whiskers showing points at 1.5$\times$~inter-quartile range.}
  \label{figure:isacosy_runtimes}
\end{figure}

\begin{figure}
  \centering
  \input{isacosyprec.pgf}
  \caption{Precision and recall of all successful IsaCoSy runs (spread
    horizontally to reduce overlapping). Lines show the combined proportion for
    each sample size (ratio of averages) and errorbars show the sample standard
    deviation.}
  \label{figure:isacosy_precRec}
\end{figure}

>>>>

If we wish to make measurable progress on the task of conjecture generation, we
must quantify what it is we are trying to achieve. Various approaches can then
be measured and compared to optimise toward our goal. To this end we define a
benchmarking methodology, shown in Figure~\ref{figure:flow_chart} on
page~\pageref{figure:flow_chart}, which both generates a large
definition/ground-truth corpus, and provides a scalable, statistical approach to
evaluating MTE tools using this corpus. We follow a precision/recall approach
similar to prior work, with the main difference being the source of definitions
and ground truths: we take existing problem sets designed for automated theorem
proving, and adapt their content for use in the conjecture generation setting.

@@

\begin{itemize}
\item For each problem, there should be a clear distinction between the
  statement to be proved and the definitions involved, such that the two can be
  easily and meaningfully separated. This rules out problem sets like those of
  SMT-COMP~\cite{barrett2005smt}, where many problems involve uninterpreted
  functions, whose behaviour is \emph{implicit} in the logical structure of the
  theorem statement but not separately \emph{defined}.
\item Definitions should be translatable into a form suitable for the MTE tools
  under study. Our application in $\S$\ref{sec:application} requires Haskell and
  Isabelle translations, and also benefits from having definitions be strongly
  typed.
\item The problem set should be relevant to the desired domain. Our focus on
  functional programming requires higher-order functions and inductively defined
  datatypes, which rules out first-order languages/logics (such as
  TPTP~\cite{sutcliffe2009tptp}).
\item The problem set should ideally contain \emph{every} ``interesting''
  property involving its included definitions, since non-membership in the
  ground truth will be treated as being ``uninteresting''. More realistically,
  we should aim for each definition to have \emph{multiple} properties; rather
  than a mixture of unrelated problems, cherry-picked from different domains.
\item Larger problem sets are preferable as they give more robust statistics, all
  else being equal (i.e. when this does not sacrifice quality).
\end{itemize}

@@

We can summarise a set of runtimes by choosing the median, as this is more
robust than the mean against long-running outliers, and hence represents
performance for a ``typical'' input of that size. Since our methodology measures
performance over many samples, we can also compute the \emph{spread} of our
data, for example the inter-quartile range. This is not possible using the
one-off measurements common in prior evaluation methods.

Considering precision and recall separately, there are two ways to average these
for a set of samples. Since both of these measurements are ratios, their means
are known as the \emph{average of the ratios} (AOR). Using the definitions of
precision and recall given in $\S$\ref{sec:te}, we can define the average of the
ratios for $S$ samples indexed by $1 \leq i \leq S$ as follows:

\begin{align*}
       \overline{\text{precision}}_{AOR}
    &= \frac{1}{S} \sum_i{\text{precision}_i} \\
    &= \frac{1}{S} \sum_i{
         \frac{\#\text{interesting}_i}
              {\#\text{generated}_i}}         \\[10pt]
       \overline{\text{recall}}_{AOR}
    &= \frac{1}{S} \sum_i{\text{recall}_i} \\
    &= \frac{1}{S} \sum_i{
         \frac{\#\text{interesting}_i}
              {\#\text{groundtruth}_i}}
\end{align*}

We interpret these mean values as the \emph{expected} precision and recall for
samples of this size. In particular, these tell us the quality of the \emph{set}
of conjectures that will be generated if we were to run the tool on such a
sample, but they don't give us direct information about the individual
conjectures themselves. Since these are mean values, they are the reference
point for measures of spread such as the variance, standard deviation and mean
absolute deviation.

The alternative is to calculate \emph{pooled} averages, by combining the
(multi)sets from each sample before taking the ratios. The size of these pooled
sets is simply the sum of the individual set sizes, and ratios of these sums are
equal to ratios of the means (since the normalising factor $\frac{1}{S}$ appears
in both numerator and denominator). Hence these values are known as the
\emph{ratio of the averages} (ROA):

\begin{align*}
       \overline{\text{precision}}_{ROA}
    &= \frac{\sum_i{\#\text{interesting}_i}}
            {\sum_i{\#\text{generated}_i}}                              \\
    &= \frac{\frac{1}{S} \sum_i{\#\text{interesting}_i}}
            {\frac{1}{S} \sum_i{\#\text{generated}_i}}                  \\
    &= \frac{\hspace{5pt} \overline{\#\text{interesting}} \hspace{5pt}}
            {\hspace{5pt} \overline{\#\text{generated}}   \hspace{5pt}} \\[10pt]
       \overline{\text{recall}}_{ROA}
    &= \frac{\sum_i{\#\text{interesting}_i}}
            {\sum_i{\#\text{groundtruth}_i}}                            \\
    &= \frac{\frac{1}{S} \sum_i{\#\text{interesting}_i}}
            {\frac{1}{S} \sum_i{\#\text{groundtruth}_i}}                \\
    &= \frac{\hspace{5pt} \overline{\#\text{interesting}} \hspace{5pt}}
            {\hspace{5pt} \overline{\#\text{groundtruth}} \hspace{5pt}}
\end{align*}

Larger sets contribute more to a ratio of averages, unlike the mean which treats
each set equally. This weighting gives us expected qualities of
\emph{conjectures} rather than sets: the ROA for precision is the expected
chance that some particular generated conjecture will appear in the ground
truth; the ROA for recall is the expected chance that some particular ground
truth property will appear in the generated output.

These averages can differ when there is a large variation in set size. For
example, if we generate a single, interesting conjecture for one sample and 99
uninteresting conjectures for another sample, the AOR (mean) precision will be
$\frac{1}{2}$ but the ROA precision will be $\frac{1}{100}$. This highlights the
importance of knowing \emph{absolute} numbers of conjectures, such as the mean
output size $\overline{\#\text{generated}}$ and its spread, alongside the
proportions given by precision and recall.

It is also possible to combine precision values \emph{with} recall values
(either individually or aggregated) by calculating their \emph{F-score}, defined
as:

\begin{equation*}
  F = 2 \times \frac{\text{precision} \times \text{recall}}
                    {\text{precision} +      \text{recall}}
\end{equation*}

This is the harmonic mean of precision and recall, ranging between 0 and 1. The
F-score summarises performance into a single number, which is useful for
purposes like ranking, but it obscures insights we might gain from looking at
precision and recall separately (e.g. whether a tool is generating too much
output or too little).

@@

To measure progress and encourage competition in the field, it is important to
compare the relative performance of different tools on the same task. Since the
aggregate statistics in the above summaries have obscured the details of
specific runs, any comparison based on them (such as comparing mean recall)
would have very low statistical power. More direct comparisons can be made using
\emph{paired} tests, since for each individual sample we have measurements for
\emph{both} tools.

Running times do not follow a normal distribution, in particular due to the
lower bound at 0, which complicates any application of standard comparisons like
the paired z-test. An alternative which does not assume normality is the
Wilcoxon signed-rank test~\cite{wilcoxon1945individual}, which has been applied
to software benchmarking in the Speedup-Test protocol of Touati, Worms and
Briais~\cite{touati2013speedup}. Our methodology differs by measuring with a
different sample each time, rather than repeatedly measuring one sample of each
size; this also allows us to use the paired form of the Wilcoxon test.

Another complication with our time measurements is the censoring effect caused
by timeouts. Paired difference tests are robust to this, since the upper-bound
on total time imposes an upper-bound to the observable difference; this biases
our results in a conservative way, \emph{towards} the null hypothesis of
indistinguishable performance.

Quality can be compared using precision and recall, but we must choose how to
handle samples where one tool succeeded and the other failed (e.g. timing out).
One approach would be to consider failed runs as if they generated an empty set
of conjectures; this conflates quality with failure, which may be desirable, but
we prefer the more charitable approach of comparing only those samples where
both tools finished successfully. We pool the results from such samples together
(allowing duplicates) and compare the resulting proportions.

Recall relies on a fixed set of properties, the ground truth, so we can compare
tools using McNemar's test for paired samples~\cite{mcnemar1947note}. For each
tool, we count how many of the ground-truth properties it found which were
\emph{not} found by the other tool. McNemar's test determines whether we can
reject the null hypothesis that both tools have the same count (i.e. both tools
have a similar contribution to the symmetric difference of their interesting
outputs).

Comparing precision cannot be done in the same way as recall, since we do not
have a fixed set of properties to divide up (there are an unlimited number of
properties that may or may not appear in a tool's output). Instead, we can count
how many conjectures were interesting and uninteresting for each tool and use
Boschloo's form of Barnard's test~\cite{lydersen2009recommended} to determine if
these proportions are independent. This assumes a binomial distribution and is
conditioned on the number of generated conjectures (as if this were fixed by
design).

@@

TIP comes with software to translate its definitions into Haskell and Isabelle
code, including comparison functions and random data generators suitable for
QuickCheck. We translate all 269 unique definitions into a single module/theory
which is imported on each run of the tools, although only those functions which
appear in the current sample are included in the signature and explored. We also
encode all names in hexadecimal to avoid problems with language-specific naming
rules, for example \texttt{add} becomes \texttt{global616464} (the prefix
distinguishes these from local variables and prevents names from beginning with
a digit). This ensures that the generated conjectures will be using the same
names as the ground truth, rather than some language-specific variant.

@@

\begin{figure}
  \centering
  \input{quickspectime.pgf}
  \caption{Running times of QuickSpec on theories sampled from TIP. Each point
    is a successful run (spread out horizontally to reduce overlaps). Red
    crosses show runs which timed out, which occurred for every size. Each size
    has 31 runs total, except for size 1 (14 runs) and size 2 (30 runs) due to
    sampling restrictions. Box plots show inter-quartile range, which reaches
    the timeout for sizes above 8. Medians are marked with a line and remain
    near zero until size 10, with those of size 13 and 20 timing out. Whiskers
    show 1.5$\times$~inter-quartile range.}
  \label{figure:quickspec_runtimes}
\end{figure}

\begin{figure}
  \centering
  \input{quickspecprec.pgf}
  \caption{Precision and recall of successful QuickSpec runs (spread
    horizontally to reduce overlapping). Lines show the mean proportion for each
    sample size (average of the ratios) and errorbars show the sample standard
    deviation. Mean precision steadily reduces from around $0.5$ at size 1 to
    around $0.1$ for size 20, whilst mean recall remains roughly flat between
    $0.2$ and $0.5$"}
  \label{figure:quickspec_precRec}
\end{figure}

@@

\begin{figure}
  \centering
  \input{isacosytime.pgf}
  \caption{Running times of IsaCoSy on theories sampled from TIP. Each point
    is a successful run (spread out horizontally to reduce overlaps). Red
    crosses are failed runs, caused by timeouts or out-of-memory. All runs of
    size 1 succeeded, whilst nothing succeeded above size 6. Each size has 31
    runs total, except for size 1 (14 runs) and size 2 (30 runs) due to sampling
    restrictions. Box plots show inter-quartile range, which grows with size,
    and lines mark the median time, which increases rapidly from around 100
    seconds at size 1 to timing out at size 4 and above. Whiskers show
    1.5$\times$~inter-quartile range.}
  \label{figure:isacosy_runtimes}
\end{figure}

\begin{figure}
  \centering
  \input{isacosyprec.pgf}
  \caption{Precision and recall of all successful IsaCoSy runs (spread
    horizontally to reduce overlapping; all runs above size 6 failed). Lines
    show the mean proportion for each sample size (average of the ratios), which
    trends downwards for precision and appears flat for recall, although there
    are too few datapoints to be definitive. Errorbars show the sample standard
    deviation.}
  \label{figure:isacosy_precRec}
\end{figure}
